---
title: "Introduction to weaiR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to weaiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

The `weaiR` package computes the Women's Empowerment in Agriculture Index
(WEAI) family of indices from pre-computed binary adequacy indicators. It
supports:

- **pro-WEAI** (3 domains, 10 indicators)
- **WEAI** (5 domains, 10 indicators)
- **A-WEAI** (5 domains, 6 indicators)
- Custom domain/indicator/weight configurations

## Quick Start

```{r basic}
library(weaiR)
data(weai_sample)

# Compute pro-WEAI
res <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex", female = 2, hhid = "hhid"
)

print(res)
```

## Understanding the Output

The result is a `weai_result` object containing:

- `$individual`: Per-person empowerment scores, parity status, and indices
- `$aggregate`: Population-level H, A, M0, DE, HGPI, IGPI, GPI, WEAI
- `$decomposition`: Per-indicator uncensored/censored headcounts and contributions

### Aggregate Results

```{r aggregate}
# Access individual components
res$aggregate$overall$WEAI
res$aggregate$de_female$DE
res$aggregate$gpi$GPI
```

### Individual-Level Data

```{r individual}
head(res$individual[, c("hhid", "sex", "emp_score", "empowered",
                         "gender_parity")])
```

### Detailed Summary

Use `summary()` to see the full decomposition tables:

```{r summary}
summary(res)
```

## Subgroup Analysis

Add `by` to decompose results by a grouping variable:

```{r subgroup}
res_by <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex", female = 2, hhid = "hhid", by = "group"
)

# Access subgroup results
names(res_by$subgroup)
res_by$subgroup[["1"]]$weai
```
