---
title: "Using Survey Designs with weaiR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Survey Designs with weaiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Survey-Weighted Estimation

When working with complex survey data, you can pass a `survey.design` object
to get properly weighted estimates with standard errors.

### Creating a Survey Design

```{r survey, eval = requireNamespace("survey", quietly = TRUE)}
library(weaiR)
library(survey)
data(weai_sample)

# Remove rows with missing indicators
complete <- complete.cases(weai_sample[, c(
  "autonomy_inc", "selfeff", "never_violence",
  "feelinputdecagr", "assetownership", "credit_accdec",
  "incomecontrol", "work_balance", "mobility", "groupmember"
)])
ws_clean <- weai_sample[complete, ]

# Create a simple survey design (SRS)
des <- svydesign(ids = ~1, data = ws_clean)

# Compute WEAI with survey design
res <- weai(
  design = des,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex", female = 2, hhid = "hhid"
)

print(res)
```

### When to Use Survey Designs

Use the `design` parameter when:

- Your data comes from a stratified or clustered sample
- You have sampling weights (probability weights)
- You need standard errors for the aggregate estimates

When `design` is provided:

- All aggregate statistics (H, A, M0, HGPI, IGPI) use `survey::svymean()`
- Standard errors are stored in the result object
- The `data` parameter is ignored (data is extracted from the design)

### Accessing Standard Errors

```{r se, eval = FALSE}
# Standard errors are in the se_ fields
res$aggregate$de_female$se_H
res$aggregate$de_female$se_M0
```
