---
title: "Technical Note: Computing the WEAI Family of Indices with weaiR"
author: "weaiR Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical Note: Computing the WEAI Family of Indices with weaiR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## 1. Introduction

The Women's Empowerment in Agriculture Index (WEAI) is a standardized,
survey-based tool developed to track gender equality and measure empowerment,
agency, and women's inclusion in the agricultural sector (Alkire et al. 2013).
Since its initial release in 2012, abbreviated (A-WEAI; Malapit et al. 2017) and
project-level (pro-WEAI; Malapit et al. 2019) versions have been developed.

The `weaiR` package provides an R implementation for computing all three versions
of the WEAI from pre-computed binary adequacy indicators. The package implements
the Alkire-Foster methodology (Alkire and Foster 2011) and produces results
equivalent to the Stata `weai` command (version 2.0.0).

This technical note describes the mathematical framework, the computational
algorithms, and practical usage of the package.

## 2. Mathematical Framework

### 2.1 Notation

Let $n$ denote the number of individuals in the sample and $d$ the total number
of binary adequacy indicators. For each individual $i$ and indicator $j$:

- $x_{ij} \in \{0, 1\}$: adequacy status (1 = adequate, 0 = inadequate)
- $w_j > 0$: indicator weight, where $\sum_{j=1}^{d} w_j = 1$
- $k \in (0, 1]$: empowerment cutoff (default: 0.80)

### 2.2 Individual-Level Scores

The **deprivation score** for individual $i$ is the weighted share of indicators
in which the individual is inadequate:

$$c_i = \sum_{j=1}^{d} w_j \cdot (1 - x_{ij})$$

The **empowerment score** is the complement:

$$e_i = 1 - c_i$$

An individual is classified as **empowered** if $e_i \geq k$, and
**disempowered** otherwise.

### 2.3 The 5DE/3DE Sub-Index

The Domains of Empowerment sub-index is based on the adjusted headcount ratio
$M_0$ from the Alkire-Foster methodology:

**Disempowerment headcount ratio** ($H$): the proportion of individuals who
are disempowered:

$$H = \frac{1}{n} \sum_{i=1}^{n} \mathbf{1}(e_i < k)$$

**Average deprivation intensity** ($A$): the mean deprivation score among
the disempowered:

$$A = \frac{\sum_{i=1}^{n} c_i \cdot \mathbf{1}(e_i < k)}{\sum_{i=1}^{n} \mathbf{1}(e_i < k)}$$

**Adjusted headcount** ($M_0$): combines prevalence and intensity:

$$M_0 = H \times A$$

The **DE sub-index** is:

$$\text{DE} = 1 - M_0$$

The sub-index is called 5DE when computed with 5 domains (WEAI, A-WEAI) and
3DE when computed with 3 domains (pro-WEAI).

### 2.4 The Gender Parity Index (GPI)

The GPI measures intrahousehold equality in empowerment. It is computed on
**dual-adult households** only---households with exactly one male and one female
respondent.

#### Step 1: Censored Inadequacy Scores

For each individual in a dual-adult household, the censored inadequacy score
sets a floor at $1 - k$:

$$\tilde{c}_i = \max(c_i,\ 1 - k)$$

This ensures that even empowered individuals receive a minimum inadequacy
score when computing parity gaps.

#### Step 2: Gender Parity Status

A household achieves **gender parity** if either:

1. The woman is empowered ($e_{\text{woman}} \geq k$), **or**
2. The woman's censored inadequacy does not exceed the man's
   ($\tilde{c}_{\text{woman}} \leq \tilde{c}_{\text{man}}$).

#### Step 3: Parity Gap

For households *without* gender parity, the normalized empowerment gap is:

$$g_h = \frac{\tilde{c}_{\text{woman}} - \tilde{c}_{\text{man}}}{1 - \tilde{c}_{\text{man}}}$$

#### Step 4: GPI Components

**Headcount ratio without parity** ($H_{\text{GPI}}$): proportion of
dual-adult households lacking gender parity.

**Mean empowerment gap** ($I_{\text{GPI}}$): average gap among households
without parity.

$$\text{GPI} = 1 - H_{\text{GPI}} \times I_{\text{GPI}}$$

### 2.5 The Overall WEAI

The WEAI is a weighted composite of the female DE sub-index and the GPI:

$$\text{WEAI} = 0.9 \times \text{DE}_{\text{female}} + 0.1 \times \text{GPI}$$

### 2.6 Indicator-Level Decomposition

The $M_0$ measure can be decomposed by indicator to identify the drivers
of disempowerment:

**Uncensored headcount** for indicator $j$: the proportion of all
individuals who are inadequate in indicator $j$:

$$h_j^{U} = \frac{1}{n} \sum_{i=1}^{n} (1 - x_{ij})$$

**Censored headcount** for indicator $j$: the proportion of individuals who
are *both* disempowered *and* inadequate in indicator $j$:

$$h_j^{C} = \frac{1}{n} \sum_{i=1}^{n} (1 - x_{ij}) \cdot \mathbf{1}(e_i < k)$$

**Absolute contribution** of indicator $j$ to $M_0$:

$$C_j = w_j \cdot h_j^{C}$$

Note that $\sum_{j} C_j = M_0$.

**Relative contribution** (percentage):

$$R_j = \frac{C_j}{M_0} \times 100\%$$

The relative contributions sum to 100%.

## 3. WEAI Variants

### 3.1 pro-WEAI (3 Domains, 10 Indicators)

| Domain | Indicator | Weight |
|:-------|:----------|:-------|
| Intrinsic agency | Autonomy in income | 1/10 |
| | Self-efficacy | 1/10 |
| | Attitudes about IPV | 1/10 |
| Instrumental agency | Input in livelihood decisions | 1/10 |
| | Ownership of land and assets | 1/10 |
| | Access to financial services | 1/10 |
| | Control over income | 1/10 |
| | Work balance | 1/10 |
| | Visiting important locations | 1/10 |
| Collective agency | Group membership | 1/10 |

Default weights: $w_j = 1/10$ for all $j$ (equal per indicator).

### 3.2 WEAI (5 Domains, 10 Indicators)

| Domain | Indicator | Weight |
|:-------|:----------|:-------|
| Production | Input in productive decisions | 1/10 |
| | Autonomy in production | 1/10 |
| Resources | Ownership of assets | 1/15 |
| | Purchase/sale/transfer of assets | 1/15 |
| | Access to and decisions about credit | 1/15 |
| Income | Control over use of income | 1/5 |
| Leadership | Group membership | 1/10 |
| | Speaking in public | 1/10 |
| Time | Workload | 1/10 |
| | Leisure | 1/10 |

Default weights: $w_j = \frac{1}{5 \times d_k}$ where $d_k$ is the number
of indicators in domain $k$ (equal weight per domain, split equally within).

### 3.3 A-WEAI (5 Domains, 6 Indicators)

| Domain | Indicator | Weight |
|:-------|:----------|:-------|
| Production | Input in productive decisions | 1/5 |
| Resources | Ownership of assets | 2/15 |
| | Access to credit | 1/15 |
| Income | Control over use of income | 1/5 |
| Time | Work balance | 1/5 |
| Leadership | Group membership | 1/5 |

Custom weights must be specified for the A-WEAI; the default 5-domain
weighting does not reproduce the published A-WEAI weights.

## 4. Usage

### 4.1 Basic Pro-WEAI

```{r proweai}
library(weaiR)
data(weai_sample)

res <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex",
  female = 2,
  hhid = "hhid"
)

print(res)
```

### 4.2 A-WEAI with Custom Weights

```{r aweai}
res_aweai <- weai(
  weai_sample,
  domains = list(
    d1 = c("feelinputdecagr"),
    d2 = c("assetownership", "credit_accdec"),
    d3 = c("incomecontrol"),
    d4 = c("work_balance"),
    d5 = c("groupmember")
  ),
  weights = list(
    d1 = 0.2,
    d2 = c(0.13333, 0.06667),
    d3 = 0.2,
    d4 = 0.2,
    d5 = 0.2
  ),
  sex = "sex",
  female = 2,
  hhid = "hhid"
)

print(res_aweai)
```

### 4.3 Changing the Empowerment Cutoff

```{r cutoff}
res_75 <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance"),
    collective = c("groupmember")
  ),
  cutoff = 0.75,
  sex = "sex",
  female = 2,
  hhid = "hhid"
)

print(res_75)
```

### 4.4 Subgroup Decomposition

```{r subgroup}
res_by <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex",
  female = 2,
  hhid = "hhid",
  by = "group"
)

# Subgroup-specific WEAI values
sapply(res_by$subgroup, function(sg) sg$weai)
```

### 4.5 Detailed Decomposition

The `summary()` method displays uncensored and censored inadequacy headcount
ratios and the relative contribution of each indicator to disempowerment:

```{r summary}
summary(res)
```

### 4.6 Visualization

The `plot()` method produces a stacked bar chart showing the absolute
contribution of each indicator to disempowerment (requires `ggplot2`):

```{r plot, fig.cap="Absolute contribution of each indicator to disempowerment."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot(res)
}
```

### 4.7 Accessing Individual-Level Results

The `$individual` component contains per-person results that can be used
for further analysis:

```{r individual}
head(res$individual[, c("hhid", "sex", "emp_score", "empowered",
                         "gender_parity", "hh_ineq")])
```

### 4.8 Survey-Weighted Estimation

When data come from a complex survey, pass a `survey.design` object to
obtain properly weighted estimates:

```{r survey, eval = requireNamespace("survey", quietly = TRUE)}
library(survey)

# Remove rows with missing indicators for clean design
complete <- complete.cases(weai_sample[, 4:13])
ws_clean <- weai_sample[complete, ]

des <- svydesign(ids = ~1, data = ws_clean)

res_svy <- weai(
  design = des,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex",
  female = 2,
  hhid = "hhid"
)

print(res_svy)
```

## 5. Data Requirements

The input data must satisfy the following conditions:

1. **Binary indicators**: Each adequacy indicator must contain only values
   0 (inadequate), 1 (adequate), or `NA` (missing). Rows with any missing
   indicator are excluded from the estimation sample.

2. **Sex variable**: Must contain exactly two unique non-missing values.
   The `female` argument specifies which value identifies female respondents.

3. **Household ID**: Must uniquely identify households. Each household
   should contain at most one male and one female respondent for the GPI
   to be computed. Households with a single respondent are used for the
   DE sub-index but excluded from the GPI.

4. **At least 3 domains**: The domain structure must specify at least
   3 groups of indicators.

5. **Weights sum to 1**: If custom weights are provided, they must sum
   to 1 (a tolerance of 0.01 is allowed, after which they are rescaled).

## 6. Computational Notes

### 6.1 Handling of Missing Values

Individuals with a missing value on any indicator are excluded from the
entire analysis. A message reports the number of excluded observations.
This matches the behavior of the Stata `weai` command.

### 6.2 Numerical Precision

Small deprivation scores (below 0.0006) are set to zero, and the
empowerment threshold is applied with a tolerance of $10^{-4}$. These
choices match the Stata implementation and ensure consistent
classification at boundary values.

### 6.3 Performance

The package uses vectorized operations throughout. The GPI computation---the
most complex step, requiring household-level matching of male and female
respondents---uses `match()` and vectorized indexing rather than
row-by-row loops. On a dataset of approximately 23,000 observations
(12,000 households), `weai()` completes in under 0.1 seconds.

## 7. Validation

The `weaiR` package produces results identical to the Stata `weai` command
(version 2.0.0, December 2025) when applied to the same data. Version 2.0.0
of the Stata command corrected a calculation error in the relative
contribution of each indicator to disempowerment, which now correctly
sums to 100%. The `weaiR` package implements the corrected formula.

## References

Alkire, S., and J. Foster. 2011. "Counting and Multidimensional Poverty
Measurement." *Journal of Public Economics* 95 (7--8): 476--87.
doi:10.1016/j.jpubeco.2010.11.006.

Alkire, S., R. Meinzen-Dick, A. Peterman, A. Quisumbing, G. Seymour,
and A. Vaz. 2013. "The Women's Empowerment in Agriculture Index."
*World Development* 52: 71--91. doi:10.1016/j.worlddev.2013.06.007.

Malapit, H., S. Kadiyala, A. Quisumbing, K. Cunningham, and P. Tyagi.
2015. "Women's Empowerment Mitigates the Negative Effects of Low
Production Diversity on Maternal and Child Nutrition in Nepal."
*Journal of Development Studies* 51 (8): 1097--123.

Malapit, H., A. Quisumbing, R. Meinzen-Dick, G. Seymour, E. Martinez,
J. Heckert, D. Rubin, A. Vaz, K. Yount, and the Gender Agriculture and
Assets Project Phase 2 (GAAP2) Study Team. 2019. "Development of the
Project-Level Women's Empowerment in Agriculture Index (pro-WEAI)."
*World Development* 122: 675--92. doi:10.1016/j.worlddev.2019.06.018.
