---
title: "WEAI Variants: pro-WEAI, WEAI, and A-WEAI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WEAI Variants: pro-WEAI, WEAI, and A-WEAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## WEAI Variants

The `weai()` function supports all major variants of the Women's Empowerment
in Agriculture Index. The variant is determined by the number of domains and
indicators you specify.

### pro-WEAI (3 domains, 10 indicators)

The project-level WEAI uses 3 domains with equal indicator weights (1/10 each):

```{r proweai}
library(weaiR)
data(weai_sample)

res_pro <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "selfeff", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance", "mobility"),
    collective = c("groupmember")
  ),
  sex = "sex", female = 2, hhid = "hhid"
)
print(res_pro)
```

### A-WEAI (5 domains, 6 indicators)

The abbreviated WEAI uses 5 domains with specific weights:

```{r aweai}
res_aweai <- weai(
  weai_sample,
  domains = list(
    d1 = c("feelinputdecagr"),
    d2 = c("assetownership", "credit_accdec"),
    d3 = c("incomecontrol"),
    d4 = c("work_balance"),
    d5 = c("groupmember")
  ),
  weights = list(
    d1 = 0.2,
    d2 = c(0.13333, 0.06667),
    d3 = 0.2,
    d4 = 0.2,
    d5 = 0.2
  ),
  sex = "sex", female = 2, hhid = "hhid"
)
print(res_aweai)
```

### Custom Cutoff

You can change the empowerment threshold (default is 0.80):

```{r cutoff}
res_75 <- weai(
  weai_sample,
  domains = list(
    intrinsic = c("autonomy_inc", "never_violence"),
    instrumental = c("feelinputdecagr", "assetownership", "credit_accdec",
                     "incomecontrol", "work_balance"),
    collective = c("groupmember")
  ),
  cutoff = 0.75,
  sex = "sex", female = 2, hhid = "hhid"
)
print(res_75)
```

## Default Weights

When `weights` is not provided:

- **3 domains**: Equal weight per indicator (1/N for each indicator)
- **5 domains**: Equal weight per domain (1/5), split equally within domains
